---
title: "ggplot lesson 6 recap"
format: html
editor: visual
---

## ggplot recap exercises

**Load the tidyverse**

```{r}
library(tidyverse)
```

These exercises are all based on the `mpg` dataset we played with during our last class.\
Remember you can learn about the variables by typing `?mpg`

1.  **Why is this code not generating a plot? Can you fix it.**

```{r}
ggplot(data = mpg)
+ geom_point(mapping = aes(x = displ, y = hwy))
```

2.  **Change the size of the displayed datapoints in this plot**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

3.  **Here's another plot. What is the problem here? Can you fix it? And then can you make the trend-line thicker**?

```{r}
ggplot(mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth()

```

4.  **Can you split this into separate plots for cars with different number of cylinders?**

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class))

```
