---
title: "ggplot recap"
date: "2024-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## GGplot recap exercises


Load the tidyverse
```{r}
library(tidyverse)
```


These exercises are all based on the `mpg` dataset we played with during our last class.  
Remember you can learn about the variables by typing `?mpg`

<br>


Why is this code not generating a plot? Can you fix it.
```{r}
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy))
```



Change the size of the displayed datapoints in this plot

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```



Here's another plot. What is the problem here? Can you fix it? And then can you make the trendline thicker?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  geom_smooth()

```



Here's another plot. What is the problem here? Can you fix it? And then can you make the trend line thicker?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  geom_smooth()

```


Can you split this into separate plots for cars with different number of cylinders?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class))
```
 

