---
title: "Lesson 16: Functions"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

```



## Final Presentations

Please indicate whether you will be giving a final class presentation next Thursday 

* Show us how you either already have or could think of applying any part of the course material to your own work
* Be as specific as possible
* Very short ~ 1.5 mins
* Required of students taking the course for credit
* Auditors are highly encouraged to share as well!


## Individual Project Consultations

This week has our last regular lab sessions. Next week, Azwad will instead offer individual project consultations. Sign up if you're interested.


## Today's learning objectives
Today, we will cover a brief introduction to how to write your own functions in R.

By the end of today's class, you should be able to:

* Write a simple function to automate a task
* Set default values for function arguments
* Explain why we should divide code into small, single-purpose functions


## Exercise 
We've talked about how we can change the file type that `ggsave()` will output just by changing the extension of the specified name we want to give the file. It works like this:

```{r, eval = FALSE}

# To save a .png file
ggsave(filename = "figures/Europe/Germany_gdpPercap.png", plot = my_plot)
# To save a .jpg file
ggsave(filename = "figures/Europe/Germany_gdpPercap.jpg", plot = my_plot)
# To save a .pdf file
ggsave(filename = "figures/Europe/Germany_gdpPercap.pdf", plot = my_plot)

```


**Your task:** Add an argument to our function that specifies the file type you want for the plot and edit the function so that it will output the requested file type.

If you have more time, you can also add an additional argument that specifies the plot type (x-y scatter, line plot etc) and adjust the function to accommodate this.
