---
title: "Lesson 15: Iteration Part 2"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

```


## Assignments

No assignment due this week, but we have covered most of the material for Assignment 8, so you may want to start that and submit early.


## Final Presentations

Please indicate whether you will be giving a final class presentation next Thursday 

* Show us how you either already have or could think of applying any part of the course material to your own work
* Be as specific as possible
* Very short ~ 1.5 mins
* Required of students taking the course for credit
* Auditors are highly encouraged to share as well!



## Learning objectives for today

By the end of today's class, you should be able to:

* Write a `for` loop to repeat operations on different input
* Implement `if` and `if else` statements for conditional execution of code


## Code duplication

Three main benefits:

* It’s easier to see the intent of your code, because your eyes are drawn to what’s different, not what stays the same.

* It’s easier to respond to changes in requirements. As your needs change, you only need to make changes in one place, rather than remembering to change every place that you copied-and-pasted the code. You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

* You’re likely to have fewer bugs because each line of code is used in more places.


## The basic structure of a `for` loop is:
```{r, eval=FALSE}
for (each_item in set_of_items) {
  do a thing
}
```


## Exercise

Start with the for loop we developed last class

1. Modify our `for` loop so that it: 
    - loops through countries in Europe only
    - plots the product of gdp_per_cap and population size per year (should approximate the total GDP) instead of the gdp_per_cap
    - saves the plots to a new subfolder inside the (recreated) figures folder called "Europe".
1. Sync to GitHub


## .gitignore files

Lists files that git should ignore

<br>

Can be specific files, entire folders, or file types

Can use `*` as wildcards, e.g. `plots/*.pdf`



## if statement basic structure

```{r, eval=FALSE}
# if
if (condition is true) {
  do something
}

# if, else
if (condition is true) {
  do something
} else {  # that is, if the condition is false,
  do something different
}
```


## ifelse() function

```{r, eval = FALSE}
ifelse(condition is true, perform action, perform alternative action)
```


## Exercise

Compute the mean of every column in the dataset `mtcars` (comes with R)