---
title: "Lesson 15: Iteration Part 1"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

```


## End of course presentations

Thursday Oct 30 during class

Give a **very brief** presentation showing the rest of the class either

* One or several ways in which you have already implemented something from the class in your own work, or

* An overview of changes you plan to make in your workflow moving forward

<br>

Please indicate if you will be giving a presentation in the google doc


## Updated assignment 8 posted today

If you have already made substantial progress on this assignment based on the previously posted version, please let us know





## Learning objectives 

Today, we'll begin exploring the key programming concepts of iteration and conditional execution. By the end of today's and next Tuesday's class, you should be able to:

* Write a `for` loop to repeat operations on different input
* Implement `if` and `if else` statements for conditional execution of code



## Code duplication

Three main benefits:

* It’s easier to see the intent of your code, because your eyes are drawn to what’s different, not what stays the same.

* It’s easier to respond to changes in requirements. As your needs change, you only need to make changes in one place, rather than remembering to change every place that you copied-and-pasted the code. You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

* You’re likely to have fewer bugs because each line of code is used in more places.


## The basic structure of a `for` loop is:

```{r, eval=FALSE, include = TRUE}
for (each_item in set_of_items) {
  do a thing
}
```


## Exercise

Use the worksheet we copied from the cloned GitHub repo to

1. Modify our `for` loop so that it: 
    - loops through countries in Europe only
    - plots the product of gdp_per_cap and population size per year (should approximate the total GDP) instead of the gdp_per_cap
    - saves the plots to a new subfolder inside the (recreated) figures folder called "Europe".
1. Sync to GitHub




## if statement basic structure

```{r, eval=FALSE}
# if
if (condition is true) {
  do something
}

# if, else
if (condition is true) {
  do something
} else {  # that is, if the condition is false,
  do something different
}
```

