---
title: "Lesson 14: Iteration Part 2"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

```


## Next week

Tuesday: No class (Fall break)

Thursday: Guest lecture by Erika Mudrak, CSCU




## Assignments

Assignment 6 due today


## When submitting assignments

* Make sure your RMarkdown "output type" is set to "github_document" (this will create an .md file when you knit rather than .html)
* Do not put `echo=FALSE` , `include=FALSE`, or `eval=FALSE` in your exercise code chunks. These will tell R to either not show your code, not show your plots, or not run the code at all. We need to see your code AND output for grading
* Knit your RMarkdown document - this will create the .md file
* Commit/push the RMarkdown (.Rmd), the knitted file (.md), AND the assignment "files" folder (this consists of the plot images - your plots will not appear in your knitted document if you do not push this folder)



## Learning objectives for today

By the end of today's class, you should be able to:

* Write a `for` loop to repeat operations on different input
* Implement `if` and `if else` statements for conditional execution of code


## Code duplication

Three main benefits:

* It’s easier to see the intent of your code, because your eyes are drawn to what’s different, not what stays the same.

* It’s easier to respond to changes in requirements. As your needs change, you only need to make changes in one place, rather than remembering to change every place that you copied-and-pasted the code. You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

* You’re likely to have fewer bugs because each line of code is used in more places.


## The basic structure of a `for` loop is:
```{r, eval=FALSE}
for (each_item in set_of_items) {
  do a thing
}
```


## .gitignore files

Lists files that git should ignore

<br>

Can be specific files, entire folders, or file types

Can use `*` as wildcards, e.g. `plots/*.pdf`



## if statement basic structure

```{r, eval=FALSE}
# if
if (condition is true) {
  do something
}

# if, else
if (condition is true) {
  do something
} else {  # that is, if the condition is false,
  do something different
}
```


## ifelse() function

```{r, eval = FALSE}
ifelse(condition is true, perform action, perform alternative action)
```



## Exercise

Write a for loop that for each year creates a separate .csv file with the gapminder data for all countries for that year


## Exercise

You have a vector

```{r, eval = FALSE}
mass_lbs <- c(2.2, 3.5, 9.6, 1.2)
```

First, write a for loop that loops over this vector and prints out the mass in kilograms 

```{r, eval = FALSE}
mass_kg = 2.2 * mass_lbs
```

Second, write a for loop that stores the calculated mass in kg values in a second vector



