---
title: "Lesson 9: ggplot Part 2"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

```

```{r}
library(tidyverse)
coronavirus <- read_csv('https://raw.githubusercontent.com/RamiKrispin/coronavirus/master/csv/coronavirus.csv')

```



## Learning objectives 

By the end of this class, you should be able to:

* Use different geometric objects and aesthetics to explore various plot types in ggplot
* Chain together steps for data wrangling (dplyr) and plotting (ggplot)


## Exercise

Reproduce this plot showing the history of daily counts of new confirmed cases worldwide (we need the total count of confirmed cases per day)

```{r, echo = FALSE}
coronavirus %>%  
  filter(type == "confirmed") %>% 
  group_by(date) %>% 
  summarize(total_cases = sum(cases)) %>% 
  ggplot(mapping = aes(x = date, y = total_cases)) +
  geom_line()
```


## Exercise

Now generate a plot with a separate line showing the daily reports of new confirmed cases in each country

