---
title: "Lesson 13: Wrap-up on relational data, then on to Factors"
format:
  revealjs:
    code-line-numbers: false
editor: visual
execute:
  eval: false
  echo: true
---

## Announcements

<br>

::: incremental
- **Upcoming classes:**
  - Thursday 10/6: Azwad will provide an overview of using AI in coding (responsibly)
  - Tuesday 10/13: No class (Feb break)
  - Thursday 10/5: Erika Mudrak from the Cornell Statistical Service Unit will visit us to share tips on troubleshooting and getting help
:::

## Today's learning objectives

<br>

Wrap-up on join and filter functions (**relational data**)

<br>

Then on to **Factors**.

- By the end of today's class, you should be able to:

  - Describe key features of factor variables in R

  - Manipulate factor levels to improve plots of categorical data

## Recap: relational data

<br>

Relational data stores information across **multiple tables** that are connected by **keys**.

Before joining two tables, ask:

1.  What does **one row** represent in each table?
2.  What **key** connects them?
3.  Is the key **unique** in either table?
4.  Which observations do I want to keep?

## Primary and foreign keys

<br>

```{r}
planes |> select(tailnum, manufacturer, model)

flights |> select(tailnum, flight, origin, dest)
```

<br>

**Primary key:** uniquely identifies an observation in its own table.

**Foreign key:** refers to the primary key of another table.

Here:

- `planes$tailnum` = **primary key**
- `flights$tailnum` = **foreign key**

::: callout-note
Primary and foreign describe the **role of a key within a table** — not which table is on the left or right of a join.
:::

## Think before you join

<br>

```{r}
flights |>
  left_join(planes, by = "tailnum")
```

<br>

How would you describe what this does in words?

<br>

. . .

> **Keep every flight and look up information about the plane used for that flight.**

What do you expect to happen to the number of rows?

## Many-to-one relationships

<br>

Suppose we want to add information about the plane to each flight.

:::::: columns
::: {.column width="42%"}
**`flights`**

| flight | tailnum |
|-------:|:--------|
|   1545 | N123    |
|   1714 | N123    |
|   1141 | N456    |
:::

::: {.column width="16%"}
<br><br>

**→**
:::

::: {.column width="42%"}
**`planes`**

| tailnum | seats |
|:--------|------:|
| N123    |   149 |
| N456    |   186 |
:::
::::::

<br>

- One plane can appear in **many flights**
- Each `tailnum` appears **once in `planes`**
- This is a **many-to-one** relationship

```{r}
flights |>
  left_join(planes, by = "tailnum")
```

**What should happen to the number of rows?**

## The many-to-many problem

Now imagine `tailnum` is **not unique in either table**.

:::::: columns
::: {.column width="45%"}
**`flights`**

| flight | tailnum |
|-------:|:--------|
|   1545 | N123    |
|   1714 | N123    |
|   1141 | N456    |
:::

::: {.column width="10%"}
<br><br>

**×**
:::

::: {.column width="45%"}
**`plane_info`**

| tailnum | status |
|:--------|:-------|
| N123    | A      |
| N123    | B      |
| N456    | A      |
:::
::::::

**What happens when we join by `tailnum`?**

## The many-to-many problem

For `N123`, there are **2 matches in each table**:

| flight | tailnum | status |
|-------:|:--------|:-------|
|   1545 | N123    | A      |
|   1545 | N123    | B      |
|   1714 | N123    | A      |
|   1714 | N123    | B      |

So:

$$2 \times 2 = 4 \text{ rows}$$

::: callout-warning
If a join unexpectedly **increases the number of rows**, check whether your key is unique in the table where you expected one match.
:::

## Quick check: which join?

<br>

Starting with `flights`, what would you use if you wanted to...

<br>

| Goal                                                 | Join |
|------------------------------------------------------|------|
| Add information about the plane                      |      |
| Keep only flights with plane information available   |      |
| Find flights with **no** plane information available |      |

<br>

## Quick check: which join?

<br>

Starting with `flights`, what would you use if you wanted to...

<br>

| Goal                                                 | Join          |
|------------------------------------------------------|---------------|
| Add information about the plane                      | `left_join()` |
| Keep only flights with plane information available   | `semi_join()` |
| Find flights with **no** plane information available | `anti_join()` |

<br>

## Your turn

<br>

We want to add the **manufacturer** and **number of seats** of the aircraft to every flight.

<br>

Before writing any code:

1.  What tables do we need?
2.  What is the key?
3.  Where is it a primary key?
4.  Where is it a foreign key?
5.  What join should we use?
6.  What should happen to the number of rows?

. . .

```{r}
flights |>
  left_join(
    planes |> select(tailnum, manufacturer, seats),
    by = "tailnum"
  )
```

## Factors

<br>

- Categorical variables
- Have a fixed and known set of possible values
- Levels are ordered
- Have human readable names, but under the hood, R is really storing integer codes 1, 2, 3, etc.
- Useful for modeling, but also when you want to display character vectors in a non-alphabetical order

## Exercise

<br>

Exercise: Filter the gapminder data down to rows where population is less than a quarter of a million, i.e. 250,000. Get rid of the unused factor levels for country and continent in different ways, such as:

- `droplevels()`
- `fct_drop()` inside `mutate()`

## Exercise

<br>

Isolate the data for "Australia", "Korea, Dem. Rep.", and "Korea, Rep." in the 2000x. Revalue the country factor levels to "Oz", "North Korea", and "South Korea"

## Exercise

<br>

Make a plot of gdpPercap for each Asian country in 2007

Make sure the countries appear in order in increasing gdpPercap

OPTIONAL: Move Thailand to be the first country displayed

## fct\_\*() functions covered today

<br>

- fct_infreq()
- fct_rev()
- fct_reorder()
- fct_reorder2()
- fct_relevel()
- fct_recode()
