---
title: "Lesson 14: Iteration Part 1"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

```


## Change of course schedule


## Assignments

* Assignment 7 due this Thursday

* Updated assignment schedule



## Learning objectives 

We'll start by working through a few exercises on using integrating data from multiple tibbles (using `join()` functions) and recap on strategies for successfully integrating relational data. Then we'll shift gears to begin exploring the key programming concepts of iteration and conditional execution.

By the end of today's and next Tuesday's class, you should be able to:

* Write a `for` loop to repeat operations on different input
* Implement `if` and `if else` statements for conditional execution of code


## Exercise

Add the infant mortality and fertility data to our original gapminder data. 

Run 
```{r, eval = FALSE, echo = TRUE}
gap_dslabs_caps <- gap_dslabs %>% 
  rename("Country" = country, "Year" = year)
```

<br>

Now add the infant mortality and fertility from the gap_dslabs_caps to our original `gapminder data

<br>

Do we have those statistics for all observations? What is a good way to figure out which countries we don't have mortality and fertility data for?


## Code duplication

Three main benefits:

* It’s easier to see the intent of your code, because your eyes are drawn to what’s different, not what stays the same.

* It’s easier to respond to changes in requirements. As your needs change, you only need to make changes in one place, rather than remembering to change every place that you copied-and-pasted the code. You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

* You’re likely to have fewer bugs because each line of code is used in more places.


## The basic structure of a `for` loop is:

```{r, eval=FALSE}
for (each_item in set_of_items) {
  do a thing
}
```


### if statement basic structure

```{r, eval=FALSE}
# if
if (condition is true) {
  do something
}

# if, else
if (condition is true) {
  do something
} else {  # that is, if the condition is false,
  do something different
}
```

