---
title: "Assignment 5: Data transformation and visualization - Part 2"
format: gfm
execute:
  echo: true
---

## Instructions: Please read through this before you begin

- This assignment is due by **8pm on Friday 10/02/2026**. Please submit it using your personal GitHub repository for this class.
- **Download this Quarto (`.qmd`) file**, place it in your personal GitHub repository for the class, and **work directly in this file**.
- Please name the file `assignment_5.qmd`. It is already set up to use `gfm` (GitHub Flavored Markdown).
- Write your code in the chunks provided and show **BOTH your code and its output**, unless instructed otherwise.
- Replace *“Write your response here”* with your response when a written answer is requested.
- Examine messages and warnings before using Quarto options to hide them.
- When finished, render `assignment_5.qmd` to produce `assignment_5.md`, and commit and push **both files**.
- Questions 2.8–2.11 are **optional**.

### Use of AI

This assignment has two goals:

1.  Practice and reinforce the data transformation and visualization skills we have been developing in class.
2.  Practice using generative AI as a **coding partner without outsourcing your understanding**.

Some questions are marked **Code yourself**. Write these without generative AI; course notes, readings, R help (`?function_name`), and package documentation are fine.

Questions marked **Use AI intentionally** ask you to document part of your AI interaction. The goal is not simply to get working code: practice writing useful prompts, evaluating responses, testing code, and deciding whether a suggestion is appropriate.

Questions marked **Evaluate AI** ask you to inspect AI-generated code or compare it with your own approach. Use your knowledge of R to evaluate the output rather than assuming it is correct or optimal.

For every AI question, **you remain responsible for the final code you submit** and should be able to explain every line.

------------------------------------------------------------------------

<br>

# Load packages

First, let's load the packages we need. You will need to install the `gapminder` package if you don't have it already.

```{r}
#| message: false

library(tidyverse)
library(gapminder) #install.packages("gapminder")
```

<br>

# Exercise 1. Trends in land value

This exercise uses a dataset describing trends in land value (`Land.Value`) in different U.S. states from 1975–2013. States are grouped into four regions under `region`.

```{r}
housing <- read_csv("https://raw.githubusercontent.com/nt246/NTRES-6100-data-science/master/datasets/landdata_states.csv")
head(housing)
```

<br>

## 1.1 Update, select, and filter — **Code yourself**

Washington, DC was not assigned to a region. According to the U.S. Census Bureau, DC is part of the South region.

Using a pipeline:

1.  Change the `region` for DC to `"South"` (`mutate()` and `ifelse()` may be helpful)
2.  Keep only `region`, `State`, `Date`, and `Land.Value`.
3.  Save the result as a new object with an informative name.
4.  Pull out the DC records and show the first six rows.

```{r}
# Write your code here
```

How many records are there for DC?

*Write your response here.*

<br>

<br>

## 1.2 Summarize by region and time — **Code yourself**

Using the updated dataset from 1.1, calculate the **mean land value for each region at each time point**. Save the result and show its first six rows.

```{r}
# Write your code here
```

Why do you need to group by **both** `region` and `Date`?

*Write your response here.*

<br>

<br>

## 1.3 Plot trends through time — **Use AI as a reviewer**

Using the summarized data from 1.2, make a line plot showing mean land value through time for each region, with regions shown in different colors.

**Write the plot yourself first, without AI.**

```{r}
# Write your code here
```

After it works, ask AI to review your code. Ask whether it accomplishes the task and to explain its reasoning **without rewriting the code unless there is an error**.

> **Paste the prompt you used:**\
> *Write your prompt here.*

Did AI identify a real problem, suggest an unnecessary change, or confirm your approach?

*Write your response here.*

<br>

<br>

# Exercise 2. Life expectancy and GDP per capita, 1952–2007

This exercise uses the `gapminder` dataset from the `gapminder` package. It contains life expectancy (`lifeExp`), GDP per capita (`gdpPercap`), and population (`pop`) for 142 countries from 1952 to 2007.

```{r}
head(gapminder)
```

<br>

## 2.1 GDP and life expectancy in 2007 — **Code yourself**

Use a scatter plot to explore the relationship between `gdpPercap` and `lifeExp` in 2007. Use both data transformation and visualization in the same pipeline.

```{r}
# Write your code here
```

<br>

## 2.2 Add a smoothing line — **Code yourself**

Add a smoothing line to the previous plot.

```{r}
# Write your code here
```

<br>

## 2.3 Compare continents — **Use AI intentionally**

Modify the plot so that it:

- excludes Oceania;
- shows each continent in a different color;
- fits a **separate smoothing line for each continent**; and
- does not show confidence intervals.

Before asking AI, write the prompt you would use. Give enough information about the data and your existing plot to get a useful answer.

> **Your AI prompt:**\
> *Write your prompt here.*

Before running the suggested code, inspect it and answer:

1.  How does the code cause `ggplot2` to fit separate smoothing lines by continent?
2.  What removes the confidence intervals?
3.  Is there anything in the suggested code that you do not understand?

*Write your response here.*

Now test it and write your final code. You may use, modify, or replace the AI solution.

```{r}
# Write your final code here
```

Did the plot accomplish all four requirements? If you changed the AI solution, explain why.

*Write your response here.*

<br>

<br>

## 2.4 Solve the same problem with facets — **Code yourself**

Show the relationship separately for each continent except Oceania using facets. Include confidence intervals.

```{r}
# Write your code here
```

<br>

<br>

## 2.5 Life expectancy through time — **Evaluate AI**

Suppose AI returned this code for: *“Show how life expectancy has changed through time in each continent, with each continent shown in a different color.”*

``` r
gapminder |>
  ggplot(aes(x = year, y = lifeExp, color = continent)) +
  geom_line()
```

**Before running it**, consider that each continent contains many countries. What do you predict? Does `geom_line()` have enough information to know which observations belong together?

*Write your response here.*

Run the AI code:

```{r}
# Paste and run the AI-generated code here
```

Now produce a more meaningful plot that still shows individual countries and colors them by continent.

```{r}
# Write your corrected code here
```

What did you change, and why does it fix the problem?

*Write your response here.*

After solving it, you may ask AI to explain the issue. Compare its explanation with yours.

*Write your response here.*

<br>

<br>

## 2.6 Investigate Asia — **Code yourself**

Create and save a dataset containing only Asian countries. Show its first six rows.

```{r}
# Write your code here
```

<br>

<br>

## 2.7 Identify unusual trends — **AI for interpretation, not observation**

Using the Asian dataset, make a plot that lets you identify countries with unusually sharp declines in life expectancy. **Facet by country.**

```{r}
# Write your code here
```

First inspect the plot yourself. Which countries show especially sharp declines, and approximately when?

**Do not use AI for this step.**

*Write your response here.*

Now choose **one** pattern and ask AI to help develop a hypothesis about a historical event that might explain it.

> **Your AI prompt:**\
> *Write your prompt here.*

What explanation did AI provide? **Do not assume it is correct.** What would you need to verify before using that explanation in a scientific analysis or publication?

*Write your response here.*

<br>

<br>

------------------------------------------------------------------------

# THE NEXT FOUR QUESTIONS ARE OPTIONAL

### If you don't want to do them, jump straight to the final recap question below (Question 3)

<br>

## 2.8 GDP per capita through time — **Code yourself** *(OPTIONAL)*

Explore the trend in per capita GDP through time on each continent.

```{r}
# Write your code here
```

<br>

## 2.9 Find an unusual GDP pattern — **Use AI to generate hypotheses** *(OPTIONAL)*

Using the Asian dataset, make a plot that identifies the country with a particularly sharp decline in GDP per capita.

```{r}
# Write your plotting code here
```

Identify the country **before using AI**.

*Write your response here.*

Ask AI for possible historical explanations.

> **Your AI prompt:**\
> *Write your prompt here.*

Which factual claims in the AI response would you want to verify with reliable sources?

*Write your response here.*

<br>

## 2.10 Compare 1952 and 2007 — **Code yourself** *(OPTIONAL)*

Starting from `gapminder`, compare `gdpPercap` and `lifeExp` in **1952 and 2007**:

- facet by year;
- color by continent;
- make point size proportional to population.

```{r}
# Write your code here
```

<br>

## 2.11 Improve a visualization — **Use AI for ideas, not code** *(OPTIONAL)*

Ask AI for **three suggestions for improving the visualization without asking it to write R code**.

> **Your AI prompt:**\
> *Write your prompt here.*

Which suggestions would improve this particular plot, and which would you reject?

*Write your response here.*

Implement the changes **yourself**.

```{r}
# Write your final plot here
```

Explain the changes you chose and why.

*Write your response here.*

<br>

# 3. Final reflection on AI and coding

1.  In which question was AI most useful, and **what kind of help** did it provide?
2.  Give one example where your own knowledge of R or visualization was necessary to evaluate an AI response.
3.  Did AI suggest anything technically correct but unnecessary or inappropriate for the task?
4.  Which part of this assignment could you now do again without AI or notes?

*Write your response here.*
