---
title: "Lab 7: Relational data and tidy data"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(base.dir = "../docs/", root.dir = "../docs/")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

<br>


<br>

## Goals for today

1. Practice connecting relational data with `join` functions in `dplyr`

2. Continue to practice data tidying with `tidyr`

3. Continue to practice data visualization with `ggplot2`
 
4. Continue to practice data transformation with `dplyr`

5. Integrate 1), 2), 3) and 4) to continue our exploration of the `babynames` dataset, and ...


<br>
<br>

## General instructions

* Today, we will combine the `join` functions and data transformation tools in `dplyr`, the data visualization tools in `ggplot2`, and the data tidying tools in `tidyr` to continue our exploration of patterns and trends in the `nycflights13` datasets we explored in class and the `babynames` dataset we worked with last week.

<br>

* To start, first open a new Quarto file in your course repo, set the output format to `github_document`, save it in your `lab` folder as `lab7.qmd`, and work in this RMarkdown file for the rest of this lab. 

<br>
<br>


## Exercise 1: Exploration of the `nycflights13` data (50 min)

We will start out with some further exploration of the datasets included in the `nycflights13` package that we worked with in Wednesday's lecture.

Let's first load in the required packages and data

```{r}
# Load required packages
library(tidyverse)
library(knitr)
library(nycflights13) # install.packages("nycflights13")

flights |> head() |> kable()
```

<br>

#### Question 1: Compute the average delay by destination, then join on the `airports` data frame so you can show the spatial distribution of delays. Here's an easy way to draw a map of the United States:


```{r, eval = FALSE}
library(maps) #install.packages("maps")

airports |>
  semi_join(flights, c("faa" = "dest")) |>
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

Don't worry if you don't understand what `semi_join()` does --- we will discuss it, or you can learn about it [here](https://r4ds.had.co.nz/relational-data.html#filtering-joins).

You might want to use the `size` or `colour` of the points to display the average delay for each airport.

<br>

*One possible solution*
<details>
  <summary>click to expand</summary>

```{r}
avg_dest_delays <-
  flights |>
  group_by(dest) |>
  # arrival delay NA's are cancelled flights
  summarise(delay = mean(arr_delay, na.rm = TRUE)) |>
  inner_join(airports, by = c(dest = "faa"))

avg_dest_delays |>
  ggplot(aes(lon, lat, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()
```

</details>

<br>
<br>

#### Question 2: Is there a relationship between the age of a plane and its delays?

Hint: Think about which of our datasets have relevant information and how we need to connect them.

<br>

*One possible solution*
<details>
  <summary>click to expand</summary>

```{r}
plane_cohorts <- inner_join(flights,
  select(planes, tailnum, year),
  by = "tailnum",
  suffix = c("_flight", "_plane")
) |>
  mutate(age = year_flight - year_plane) |>
  filter(!is.na(age)) |>
  mutate(age = if_else(age > 25, 25L, age)) |>
  group_by(age) |>
  summarise(
    dep_delay_mean = mean(dep_delay, na.rm = TRUE),
    arr_delay_mean = mean(arr_delay, na.rm = TRUE)
  )
## Departure delays
ggplot(plane_cohorts, aes(x = age, y = dep_delay_mean)) +
  geom_point() +
  scale_x_continuous("Age of plane (years)", breaks = seq(0, 30, by = 10)) +
  scale_y_continuous("Mean Departure Delay (minutes)")
## Arrival delays
ggplot(plane_cohorts, aes(x = age, y = arr_delay_mean)) +
  geom_point() +
  scale_x_continuous("Age of Plane (years)", breaks = seq(0, 30, by = 10)) +
  scale_y_continuous("Mean Arrival Delay (minutes)")
```

</details>

<br>
<br>

#### Question 3: What weather conditions make it more likely to see a delay?

Hint: Think about which of our datasets have relevant information and how we need to connect them.

<br>

*One possible solution*
<details>
  <summary>click to expand</summary>

```{r}
flight_weather <-
  flights |>
  inner_join(weather, by = c("origin", "year", "month", "day", "hour"))
## Precipitation
flight_weather |>
  group_by(precip) |>
  summarise(delay = mean(dep_delay, na.rm = TRUE)) |>
  ggplot(aes(x = precip, y = delay)) +
  geom_line() + geom_point()
# Visibility
flight_weather |>
  ungroup() |>
  mutate(visib_cat = cut_interval(visib, n = 10)) |>
  group_by(visib_cat) |>
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE)) |>
  ggplot(aes(x = visib_cat, y = dep_delay)) +
  geom_point()
```

</details>

<br>
<br>

## Recap (5 minutes)

<br>

Share your findings, challenges, and questions with the class. 

<br>
<br>

## Short break (10 min)

<br>
<br>


## Exercise 2: Baby names (45 min)

<br>

Use data tidying, transformation, and visualization to answer the following questions about baby names in breakout rooms

<br>

top boy names              | top girl names
:-------------------------:|:-------------------------:
![](https://images.squarespace-cdn.com/content/v1/5c80324b29f2cc57eb5a3e1f/1581985571871-AWAM0VICLVSKXHQZ9H8V/ke17ZwdGBToddI8pDm48kBapyj43L202fvtEGa0NsU97gQa3H78H3Y0txjaiv_0fDoOvxcdMmMKkDsyUqMSsMWxHk725yiiHCCLfrh8O1z5QPOohDIaIeljMHgDF5CVlOqpeNLcJ80NK65_fV7S1UY5275-4EsLh0HBIeyNph8KOLVRHevxluj5VQW3JRQYgm4bjm9DAHF2kOsIZRJKXnA/top+10+baby+boy+names.png?format=500w) | ![](https://images.squarespace-cdn.com/content/v1/5c80324b29f2cc57eb5a3e1f/1581985603694-1G1HBXOTRB17QEP1234D/ke17ZwdGBToddI8pDm48kBapyj43L202fvtEGa0NsU97gQa3H78H3Y0txjaiv_0fDoOvxcdMmMKkDsyUqMSsMWxHk725yiiHCCLfrh8O1z5QPOohDIaIeljMHgDF5CVlOqpeNLcJ80NK65_fV7S1UY5275-4EsLh0HBIeyNph8KOLVRHevxluj5VQW3JRQYgm4bjm9DAHF2kOsIZRJKXnA/top+10+baby+girl+names.png?format=500w)

<br>
<br>

**Instructions:**

* Load the required packages and read in the data with the following code:

<br>

```{r}
# Load required packages
library(babynames) # install.packages("babynames")

babynames |> head() |> kable()
```

<br>

* The `babynames` dataset provides the number of children of each sex given each name from 1880 to 2017 in the US. All names with more than 5 uses are included. This dataset is provided by the US Social Security Administration.

<br>

* As a reminder, to get familar with this dataset, you might want to use functions like `View()`, `dim()`, `colnames()` , and `?`. 

<br>

* Make sure that you use figures and/or tables to support your answer.

<br>

* We provide some possible solutions for each question, but we highly recommend that you **don't look at them unless you are really stuck**.

<br>
<br>

#### Question 1: What are the 6 most popular boy names and girl names of all time? How has the popularity of each of these names changed over time? This time, use the `slice_max()` function in combination with a `join` function to answer this question.

<br>

*Hint: You can start by finding the 6 most popular names for each sex in one step using `group_by()` and `slice_max()`, and then use a filtering join function to subset the original dataset.*

<br>

*One possible solution*
<details>
  <summary>click to expand</summary>

```{r}
# number of passengers in the dataset
top_6_names <- babynames |>
  group_by(sex, name) |>
  summarise(total_count=sum(n)) |>
  ungroup() |>
  group_by(sex) |>
  slice_max(order_by = total_count, n = 6)

babynames |>
  semi_join(top_6_names, by = c("sex", "name")) |>
  ggplot(aes(x=year, y=prop, group=name, color=sex)) +
  geom_line() +
  facet_wrap(~name)
```

Note: 

* `slice_max(order_by = total_count, n = 6)` selects 6 rows with the highest values in `total_count` for each unique entry in the grouping variable (in this case, males and females)

</details>

<br>
<br>

#### Question 2. Find the names that have not appeared in this dataset until 2010, but were used more than 1000 times 2010-2017 (boys and girls combined). Do you recognize any pop culture influence in these?

<br>

*Hint: You may start by creating a variable to indicate whether a row is before or after 2010.*

*Hint: `pivot_wider()` may be helpful*

*Hint: you may need to replace NAs with 0s for this exercise. `mutate()`, `ifelse()`, and `is.na()` may become handy.*

<br>

*One possible solution*
<details>
  <summary>click to expand</summary>

```{r}
new_names <- babynames |>
  mutate(threshold = ifelse(year >= 2010, "after", "before")) |>
  group_by(name, threshold) |>
  summarise(total_count = sum(n)) |>
  pivot_wider(names_from = threshold, values_from = total_count, names_prefix = "count_") |>
  mutate_all(~replace(., is.na(.), 0)) |>
  filter(count_before == 0, count_after >=1000) 

new_names |> 
  kable()

babynames |>
  filter(name %in% new_names$name) |>
  ggplot(aes(x=year, y=n, color=sex)) +
  geom_line() +
  facet_wrap(~name)
```

Note: `mutate_all(dataset, ~replace(., is.na(.), 0))` is an efficient way to replace all NAs in a dataset with 0s.

</details>

<br>
<br>

#### Question 3. Find the 12 most popular unisex names. How have the proportion of girls for each of them changed over time? 

<br>

The definition of unisex names is arbitrary, but for this exercise, let's define them as names which have proportion of girls between 10% and 90% across all time.

<br>

*Hint: You may start by summing over years in order to get a list of unisex names*

*Hint: `pivot_wider()` may be helpful*

*Hint: you may need to replace NAs with 0s for this exercise. `mutate()`, `ifelse()`, and `is.na()` may become handy.*

<br>

*One possible solution*
<details>
  <summary>click to expand</summary>

```{r fig.width=10}
unisex_names <- babynames |>
  group_by(name, sex) |>
  summarise(total_count = sum(n)) |>
  pivot_wider(names_from = sex, values_from = total_count, names_prefix = "count_") |>
  filter(!is.na(count_M), !is.na(count_F)) |>
  mutate(total_count=count_M+count_F, f_proportion = count_F / total_count) |>
  filter(f_proportion<0.9, f_proportion>0.1) |>
  arrange(-total_count)

unisex_names |>
  head(12) |>
  kable()

unisex_names |>
  head(12) |>
  ggplot(aes(x=count_M, y=count_F)) +
  ggrepel::geom_label_repel(aes(label=name)) +
  geom_point(data=unisex_names)

babynames |>
  filter(name %in% unisex_names$name[1:12]) |>
  pivot_wider(names_from = sex, values_from = c(n, prop)) |>
  mutate_all(~replace(., is.na(.), 0)) |>
  mutate(total_count=n_F+n_M, f_proportion = n_F / total_count, average_popularity = (prop_F + prop_M)/2) |>
  ggplot(aes(year, f_proportion, group=name)) +
  geom_line() +
  geom_point(aes(size = average_popularity)) +
  facet_wrap(~name) +
  ylab("proportion of girls")
```


</details>

<br>
<br>

## Recap (5 minutes)

Share your findings, challenges, and questions with the class. 

<br>
<br>



**END LAB 7**

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
